Abstract Data Type Inference

### 2015-02-28 draft.tex
Updated intro, and added related work.

  Checking that efficient implementations of concurrent objects adhere to their
  abstract data types (ADTs) can be made tractable when ADTs are given by
  symbolic, logical representations. Such representations are however difficult
  to obtain. Common practice is to design reference implementations.

  In this work we demonstrate that effective symbolic ADT representations can
  be automatically extracted from the executions of reference implementations.
  Our approach exploits key features of naturally-occurring ADTs: violations
  can be decomposed into a small set of representative patterns, and these
  patterns manifest in executions with few operations. By identifying certain
  algebraic properties of naturally-occurring ADTs, and exhaustively sampling
  executions up to a small number of operations, we extract concise symbolic
  ADT representations which are complete in practice.

>>>>

  Checking that efficient implementations of concurrent objects adhere to their
  abstract data types (ADTs) can be made tractable when ADTs are given by
  symbolic, logical representations. Obtaining these symbolic representations,
  however, is a complex task requiring rare expertise. In practice, concurrent
  objects are specified indirectly by possibly-abstract imperative reference
  implementations.

  In this work we demonstrate that effective symbolic ADT representations can
  be automatically generated from the executions of reference implementations.
  Our approach exploits key features of naturally-occurring ADTs: violations
  can be decomposed into a small set of representative patterns, and these
  patterns manifest in executions with few operations. By identifying certain
  algebraic properties of naturally-occurring ADTs, and exhaustively sampling
  executions up to a small number of operations, we generate concise symbolic
  ADT representations which are complete in practice.

### 2015-02-28 intro.tex
Updated intro, and added related work.

TODO DISCUSS THE IMPORTANCE OF OBSERVATIONAL REFINEMENT, AND THAT IT IS HARD

Recent work has demonstrated that checking observational refinement can be
performed efficiently in practice, given symbolic ADT representations,
e.g.,~using first-order logic. However, writing accurate symbolic
specifications is an expected challenge for the designers of concurrent ADTs.
Common practice is to write reference implementations. Mapping these imperative
implementations to the logical formulæ describing valid executions is a great
challenge for human developers.

In this work we demonstrate that effective symbolic ADT representations can be
extracted from the executions of reference implementations. Our approach
exploits two key features of concurrent object ADTs: that violations of each
ADT can be decomposed into a small set of representative patterns, and that
these patterns manifest in executions with few operations. The first feature
allows us to represent symbolic ADTs finitely, as exclusions of violation
patterns. The second allows us to extract violation patterns from finite
enumerations of executions.

@@

Our algebraic insight is based on grouping the operations of an execution into
\emph{matchings}. Intuitively, operations which refer to the same instances of
values belong to the same matching. For example, a {\tt pop} operation
returning the value {\tt 1} matches a preceding {\tt push(1)} operation. By
comparing executions by the characteristics of their matchings, rather than the
actual data values they use, we capture the relation among data values
independently of the data values themselves. Furthermore, the notion of
matchings provides a key algebraic property of ADTs: the executions of
naturally-occurring concurrent object ADTs are closed under the removal of
matchings. For example, any execution of the atomic stack ADT using values {\tt
1}, {\tt 2}, and {\tt 3} would remain a valid execution were all operations
using the value {\tt 2} deleted. Conversely, any execution which extends a
violating execution with additional matchings is itself a violation. This
property, along with analogous algebraic properties concerning operation order
and completion, allow us to characterize the violations of ADTs with finite
basis sets, ultimately leading to finite symbolic ADT representations.

@@

  \item Empirical validation that our ADT-inference approximation is complete
  in practice, and effective for the detection of violations to observational
  refinement~(§\ref{sec:empirical}).

>>>>

Many modern software systems rely on efficient implementations of abstract data
types (ADTs) like atomic collections, semaphores, and locks. Efficient
implementation is difficult though, because efficiency entails maximizing
concurrency, and concurrency jeopardizes atomicity. Consequently,
implementations are vulnerable to subtle and elusive bugs compromising
adherence to their ADTs~\cite{journals/cacm/staff11}, thus necessitating
automated verification. Adherence of an implementation to an ADT is captured
formally by a property called \emph{observational refinement}: any observable
behavior over the method call and return actions of the implementation is also
admitted by the the ADT.

Although checking observational refinement is a fundamentally hard
problem\footnote{See Section~\ref{sec:related} for further discussion.} for
arbitrary ADTs~\cite{journals/siamcomp/GibbonsK97, conf/esop/BouajjaniEEH13},
recent work demonstrates that naturally-occurring ADTs like atomic collections,
semaphores, and locks can be checked efficiently given effective symbolic ADT
representations~\cite{conf/popl/BouajjaniEEH15, journals/arxiv/BouajjaniEEH15,
conf/pldi/EmmiEH15}, e.g.,~specified in first-order logic. However, writing
accurate symbolic specifications is an expected challenge for ADT designers. In
practice, ADTs are specified indirectly as the observable behaviors of
possibly-abstract imperative \emph{reference implementations}. Mapping these
reference implementations to logical formulæ describing the same sets of
behaviors is a complex task requiring rare expertise.

In this work we demonstrate that effective symbolic ADT representations can be
generated automatically from the executions of reference implementations. Our
approach exploits two key features of concurrent-object ADTs: that violations
of each ADT can be decomposed into a small set of representative patterns, and
that these patterns manifest in executions with few operations. The first
feature allows us to represent symbolic ADTs finitely, as exclusions of
violation patterns. The second allows us to extract violation patterns from
finite enumerations of executions.

@@

Our algebraic insight is based on grouping the operations of an execution into
\emph{matchings}. Intuitively, operations which refer to the same instances of
values belong to the same matching. For example, a {\tt pop} operation
returning the value {\tt 1} matches a preceding {\tt push(1)} operation. By
comparing executions by the characteristics of their matchings, rather than the
actual data values they use, we capture the relation among data values
independently of the data values themselves. Furthermore, the notion of
matchings provides a key algebraic property of ADTs: the executions of
naturally-occurring concurrent object ADTs are closed under the removal of
matchings. For example, any execution of the atomic stack ADT using values {\tt
1}, {\tt 2}, and {\tt 3} would remain a valid execution were all operations
using the value {\tt 2} deleted. Conversely, any execution which extends a
violating execution with additional matchings is itself a violation. This
property, along with analogous algebraic properties concerning operation order
and completion, allow us to compare executions via a violation-preserving
embedding relation. This relation is a well-founded partial order on
executions, and thus allows us to characterize the infinite set of violations
to an ADT with a finite basis set, ultimately leading to a finite symbolic
representation.

@@

  \item Empirical validation that our ADT-inference approximation is complete
  in practice, and effective for the detection of observational-refinement
  violations~(§\ref{sec:empirical}).

### 2015-02-28 related.tex
Updated intro, and added related work.

>>>>

The idea of decomposing ADT violations into a finite set of patterns has been
proposed for atomic collections~\cite{conf/concur/HenzingerSV13,
conf/popl/DoddsHK15, conf/popl/BouajjaniEEH15}, and generalized to ADTs which
can be expressed with a particular form of recursive
definitions~\cite{journals/arxiv/BouajjaniEEH15}. These works suggest patterns
which directly recognize violations across the entire set of \emph{concurrent}
executions, effectively reducing observational refinement to the verification
of classical temporal-logic properties. Some of these patterns are expressible
only as infinite, though \emph{regular}, languages. On the contrary, in this
work we suggest patterns expressed as \emph{finite} languages, which recognize
violations in ADT \emph{kernels} — i.e.,~the sequential executions of atomic
objects, or the “concurrency-aware” executions of “concurrency-aware”
specifications~\cite{conf/podc/HemedR14}. Though by themselves these patterns
are incomplete in recognizing violations across the entire set of concurrent
executions, they are complete when considered as negative conditions on
concurrent executions’ linearization orders, effectively reducing verification
of a single concurrent execution to satisfiability
checking~\cite{conf/pldi/EmmiEH15}.
